---
title: ""
format:
  html:
    embed-resources: true
    page-layout: full
    grid:
      margin-width: 100px
author: Daniel Väisänen
editor: source
toc: true
execute:
  warning: false
  echo: true
  eval: false
---

# Description

Cohort flow / CONSORT reconciliation (referee 1.8). The final analytic N has been
reported inconsistently across the manuscript, abstract, both TARGET tables and
the title (414,537 vs 418,285 HPA participants; 1,673,136 vs 1,673,140 controls).

This script computes the exclusion cascade **in code**, from the matched cohort
through to the final analytic data set, and emits the final N as a **single
value** (`results/tables/cohort_flow.csv`). The manuscript, tables and title
should all read from this one number. Resolve the discrepancy here first —
everything else depends on it.


# Libraries
```{r}
library(tidyverse)
```


# HPA exclusion cascade (raw database → cleaned study population)

Streamlined version of the WP1 data-cleaning cascade (`dtrackr`): from the raw HPA
database to the cleaned HPA population that feeds the matching. This produces the
exact exclusion counts for the manuscript CONSORT so they reconcile with the
analytic N (the earlier text's exclusions summed to 414,537, but the analytic
cohort is 418,283 — re-run this to get the correct counts).

NOTE: reads the raw HPA + SCB files (identifiable — in-environment only). Adjust the
`XXXXXX` DNR folder name to your path, and run in a UTF-8 locale.

```{r}
#| purl: false
# (excluded from the batch pipeline: reads external identifiable files; run this
#  chunk interactively to generate the CONSORT counts + flowchart.)
library(dtrackr)

# paths — edit the DNR folder name to match your environment
hpa_dir <- here::here(
  "..",
  "..",
  "..",
  "..",
  "HPI (DNR XXXXXX)",
  "Data HPI registeruttag 2023"
)
hpa_path <- file.path(
  hpa_dir,
  "HPI",
  "HPI clean",
  "EEB_hpb_clean_2023-10-16.csv"
)
scb_path <- file.path(
  hpa_dir,
  "SCB",
  "SCB clean",
  "Grunduppgifter_2023-09-24.csv"
)
scbvars_path <- file.path(
  hpa_dir,
  "SCB",
  "Leverans",
  "Leverans_20240306",
  "with_hpi_lopnr",
  "6.organization_variables",
  "organization_data.csv"
)

# Read the files fully and let the joins carry the columns through (as in the
# original cleaning script) — LoneInk / Income_Sources_Category live in the SCB
# files but not necessarily where you'd guess, so we don't pre-select.
hpa <- read_csv(hpa_path, show_col_types = FALSE) |>
  mutate(
    SymptomBackNeck = if_else(
      SymptomBackNeck %in% c(7, 8),
      NA_real_,
      SymptomBackNeck
    )
  )

scb <- read_csv(scb_path, show_col_types = FALSE)
scb_vars <- read_csv(scbvars_path, show_col_types = FALSE)

df1 <- hpa |>
  left_join(scb, by = join_by(LopNr, Performed, Year)) |>
  left_join(scb_vars, by = join_by(LopNr, Performed, Year))

# --- locate the income variables before the exclusions ----------------------
# exclude_all() below references LoneInk and Income_Sources_Category. If either
# name collides across the joined files it is suffixed (.x / .y) and the bare
# name disappears, or it may live in a SCB file we are not reading. Report where
# each one actually is and stop with a clear message rather than dtrackr's
# cryptic "column doesn't exist" from inside exclude_all().
locate_var <- function(v) {
  in_file <- c(
    hpa = v %in% names(hpa),
    scb = v %in% names(scb),
    scb_vars = v %in% names(scb_vars)
  )
  matched <- grep(paste0("^", v, "(\\.|$)"), names(df1), value = TRUE)
  message(
    v,
    " -> in df1 as: ",
    if (length(matched)) paste(matched, collapse = ", ") else "NONE",
    " | source file(s): ",
    if (any(in_file)) {
      paste(names(in_file)[in_file], collapse = ", ")
    } else {
      "NONE"
    }
  )
  invisible(matched)
}
invisible(lapply(c("LoneInk", "Income_Sources_Category"), locate_var))

stopifnot(
  "LoneInk is not a column in df1 — see the message above and adjust the reads/joins" = "LoneInk" %in%
    names(df1),
  "Income_Sources_Category is not a column in df1 — see the message above" = "Income_Sources_Category" %in%
    names(df1)
)

options(
  dtrackr.default_message = "n={formatC(.count, big.mark = ',', format = 'fg')}"
)

hpa_cascade <- df1 |>
  # one row per person: their first HPA, up to 2021
  arrange(LopNr, Performed) |>
  mutate(hpa_number = row_number(), .by = LopNr) |>
  filter(hpa_number == 1, Year <= 2021) |>
  track(
    .headline = "First HPA per person (through 2021), n={formatC(.count, big.mark = ',')}    "
  ) |>
  # inclusion criteria as ONE step: >=2 physical AND >=2 questionnaire variables
  filter(
    rowSums(
      !is.na(pick(
        EkB_rel_VO2,
        Astrand_rel_VO2,
        HeightCM,
        WeightKG,
        BloodPressureSystolic,
        BloodPressureDiastolic
      ))
    ) >=
      2,
    rowSums(
      !is.na(pick(
        ExerciseAnswer,
        TobaccoSmoking,
        Diet,
        Health,
        StressOverall,
        SymptomBackNeck
      ))
    ) >=
      2,
    .messages = "<2 physical or <2 questionnaire variables, n={formatC(.excluded, big.mark = ',', format = 'fg')}"
  ) |>
  status("n={formatC(.count, big.mark = ',', format = 'fg')}") |>
  exclude_all(
    Year <
      1995 ~ "HPAs before 1995, n={formatC(.excluded, big.mark = ',', format = 'fg')}",
    Age <
      18 ~ "Under 18 years of age, n={formatC(.excluded, big.mark = ',', format = 'fg')}",
    LoneInk == 0 |
      is.na(
        LoneInk
      ) ~ "No verified income, n={formatC(.excluded, big.mark = ',', format = 'fg')}",
    Income_Sources_Category == "No Income Sources" |
      is.na(Income_Sources_Category) ~
      "No verified income source, n={formatC(.excluded, big.mark = ',', format = 'fg')}"
  ) |>
  comment(
    "n={formatC(.count, big.mark = ',')}"
  )

# Free the large raw objects before sourcing 01 (which reads ~130 MB and builds
# the 2M-row data_cvd) — holding df1 + the SCB reads at the same time is what
# tips memory over on the M-series Mac. hpa_cascade keeps its own copy of the
# rows it needs, so this is safe.
rm(df1, hpa, scb, scb_vars)
gc()

# Step 6 (final): matching + strict 1:4. The analytic HPA participants are those
# retained in data_cvd (built and 1:4-filtered in 01_data_source_file.R); the
# HPA in the cleaned population but NOT in data_cvd are the ones lost to matching
# / dropped as incomplete 1:4 clusters. This turns the cleaned population into
# the final analytic N in the same flow.
source(here::here("script", "01_data_source_file.R"))
analytic_hpa_ids <- data_cvd |>
  filter(treated == "HPA") |>
  pull(LopNr) |>
  unique()

hpa_cascade <- hpa_cascade |>
  filter(
    LopNr %in% analytic_hpa_ids,
    .messages = "Excluded: not matched 1:4 / incomplete cluster, n={formatC(.excluded, big.mark = ',', format = 'fg')}"
  ) |>
  ungroup(
    .headline = "Final sample, n={formatC(.count, big.mark = ',')}  "
  )

# CONSORT flowchart + the endpoint count, which should now equal the analytic N
hpa_cascade |>
  flowchart(
    filename = here::here(
      "..",
      "results",
      "figures",
      "hpa_exclusion_flowchart"
    ),
    formats = c("png", "pdf")
  )

cat(
  "Final analytic HPA participants:",
  format(nrow(hpa_cascade), big.mark = ","),
  "\n"
)
```


# Final analytic data set (the one the models actually use)
```{r}
# data_cvd is the analytic file the survival models are fitted on. Its row count,
# split by arm, IS the final analytic N.
source(here::here("script", "01_data_source_file.R"))

final_counts <- data_cvd |>
  count(treated, name = "n")

final_counts

final_hpa <- final_counts |> filter(treated == "HPA") |> pull(n)
final_control <- final_counts |> filter(treated == "Control") |> pull(n)

# 1:4 matching ratio check. 01_data_source_file.R enforces strict 1:4 (one HPA
# + four controls per cluster), so controls MUST equal 4 × HPA here. Fail loudly
# if not — the design requires complete 1:4 sets.
ratio_check <- tibble(
  hpa = final_hpa,
  control = final_control,
  expected_4x = final_hpa * 4,
  control_matches_4x = final_control == final_hpa * 4
)

ratio_check

stopifnot(
  "Analytic data is not strict 1:4 — check the cluster filter in 01_data_source_file.R" = final_control ==
    final_hpa * 4
)
```


# Exclusion cascade: matched cohort → analytic data set
```{r}
# Stage 1: the matched cohort produced by 10_propensity_score_matching.qmd
combined_matched_data <- read_csv(
  here::here("..", "data", "combined_matched_data.csv")
)

stage_matched <- combined_matched_data |>
  distinct(LopNr, .keep_all = TRUE) |>
  mutate(
    arm = if_else(
      treated == 1 | treated == "1" | treated == "HPA",
      "HPA",
      "Control"
    )
  ) |>
  count(arm, name = "n_matched")

# Stage 2: individuals carried into the register extract / CVD analytic file
stage_cvd_source <- read_csv(
  here::here("..", "data", "df_paper_3_CVD_2025-03-05.csv")
) |>
  mutate(
    arm = if_else(
      treated == 1 | treated == "1" | treated == "HPA",
      "HPA",
      "Control"
    )
  ) |>
  count(arm, name = "n_cvd_source")

# Stage 3: final analytic data set (data_cvd, after all recoding/joins in 01_)
stage_final <- data_cvd |>
  rename(arm = treated) |>
  count(arm, name = "n_final")

# assemble the cascade side by side so the drop at each stage is explicit
cohort_flow <- stage_matched |>
  full_join(stage_cvd_source, by = "arm") |>
  full_join(stage_final, by = "arm") |>
  mutate(
    dropped_matched_to_source = n_matched - n_cvd_source,
    dropped_source_to_final = n_cvd_source - n_final
  )

cohort_flow
```


# Emit the single source-of-truth N
```{r}
# one tidy file the manuscript / tables / title all read from
cohort_flow_out <- bind_rows(
  cohort_flow |>
    select(arm, n_matched, n_cvd_source, n_final),
  tibble(
    arm = "TOTAL",
    n_matched = sum(cohort_flow$n_matched, na.rm = TRUE),
    n_cvd_source = sum(cohort_flow$n_cvd_source, na.rm = TRUE),
    n_final = sum(cohort_flow$n_final, na.rm = TRUE)
  )
)

cohort_flow_out

write_csv(
  cohort_flow_out,
  here::here("..", "results", "tables", "cohort_flow.csv")
)

# headline numbers for the manuscript/title — read these, do not hand-type them
cat(sprintf(
  "FINAL N (analytic): HPA = %s, Controls = %s, Total = %s\n",
  format(final_hpa, big.mark = ","),
  format(final_control, big.mark = ","),
  format(final_hpa + final_control, big.mark = ",")
))
```
