---
title: "Untitled"
format: html
---


Translation of ICD-9 codes to ICD-10 codes is a complex task that usually requires a detailed mapping table, because there is not always an exact equivalent between the codes in these two systems. This means that it will not be an exact translation.

```{r}
library(tidyverse)
```

# read in and update ICD-convert_key

```{r}
# icd code convert key, from icd 6,7,8,9 to 10
convert_key <-
read_delim(
  here::here("..", "data", "ICD_codes_20210820.csv"),
  delim = ";",
  skip = 1,        # increase if there are lines to skip
  col_names = TRUE
)


mapped_codes <- tribble(
  ~original_icd,    ~converted_from_icd, ~n,    ~new_icd10_code,           ~plain_language_explanation,                       ~certainty,
  #------------------------------------------------------
  # ICD9
  #------------------------------------------------------
  "253",   "ICD9", 4373, "E22", "Disorders of pituitary gland (hyper/hypofunction)",           "high",
  "255",   "ICD9", 2020, "E27", "Disorders of adrenal glands",                                 "high",
  "340",   "ICD9", 1589, "G35", "Multiple sclerosis",                                          "high",
  "256",   "ICD9",  745, "E28", "Ovarian dysfunction (e.g., polycystic ovaries)",              "high",
  "252",   "ICD9",  704, "E21", "Parathyroid gland disorders (hyper-/hypoparathyroidism)",     "approx",
  "258",   "ICD9",  318, "E31", "Polyglandular endocrine disorders",                           "approx",
  "257",   "ICD9",  113, "E29", "Testicular dysfunction (e.g., hypogonadism)",                "approx",
  "254",   "ICD9",   28, "E32", "Diseases of thymus gland",                                    "approx",
  "335",   "ICD9",   23, "G12", "Anterior horn cell disease (e.g., ALS)",                      "approx",
  "234",   "ICD9",   21, "D03",  "Could be carcinoma in situ",                                 "approx",
  "148",   "ICD9",   17, "C12", "Malignant neoplasm of pyriform sinus/hypopharynx area",       "approx",
  "175",   "ICD9",    4, "C50", "Malignant neoplasm of male breast",                           "approx",
  "19",    "ICD9",    4, "C80", "Malignant neoplasm, unspecified site",                        "approx",
  
  #------------------------------------------------------
  # ICD8
  #------------------------------------------------------
  "340",   "ICD8",  522, "G35", "Multiple sclerosis",                                          "high",
  "571",   "ICD8",  344, "K74", "Chronic liver disease, cirrhosis",                             "approx",
  "147",   "ICD8",  271, "C11", "Malignant neoplasm of nasopharynx",                           "high",
  "14B10", "ICD8",  172, "C14", "	icd-10: in the range C00–C14 (lip, oral cavity, pharynx)",   "approx",
  "149",   "ICD8",   51, "C13", "Malignant neoplasm of hypopharynx",                           "high",
  "718",   "ICD8",   51, "M19", "Other arthropathies/joint disorders",                         "approx",
  "152",   "ICD8",   22, "C17", "Malignant neoplasm of small intestine",                       "approx",
  "296A",  "ICD8",   21, "F32", "Depressive episode (older code)",                              "approx",
  "342",   "ICD8",   17, "G81", "Hemiplegia (usually stroke-related)",                         "approx",
  "348",   "ICD8",   16, "G93", "Other disorders of brain",                                    "approx",
  "14C20", "ICD8",   14, "C14", "Local extension code, uncertain mapping",    "approx",
  "296C",  "ICD8",   14, "F32", "Depressive disorders (variant)",                              "approx",
  "296W",  "ICD8",   14, "F31", "Bipolar disorder (guess)",                                    "approx",
  "140",   "ICD8",   10, "C00", "Malignant neoplasm of lip",                                   "approx",
  "314",   "ICD8",   10, "F90", "Hyperkinetic disorders (ADHD)",                               "approx",
  "14B1070","ICD8",   9, "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "159",   "ICD8",   8,  "C26", "Malignant neoplasm of ill-defined digestive organs",          "approx",
  "296D",  "ICD8",   7,  "F33", "Recurrent depressive disorder",                               "approx",
  "14B1050","ICD8",  6,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14B107","ICD8",   6,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "144",   "ICD8",   5,  "C14", "Malignant neoplasm of other oral cavity sites",               "approx",
  "156",   "ICD8",   5,  "C23", "Malignant neoplasm of gallbladder",                           "approx",
  "14B05", "ICD8",   4,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14B105","ICD8",   3,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14C2005","ICD8",  3,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14C1510","ICD8",  2,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "15A25", "ICD8",   2,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "15A35", "ICD8",   2,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "15D05", "ICD8",   2,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "301X",  "ICD8",   2,  "F60", "Personality disorders",                                       "approx",
  "14B05 0","ICD8",  1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14C05", "ICD8",   1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14C0520","ICD8",  1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14C15", "ICD8",   1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "14C1505","ICD8",  1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "15A05", "ICD8",   1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "15A10", "ICD8",   1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "15D15", "ICD8",   1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "30K05", "ICD8",   1,  "check local registry", "Local extension code, uncertain mapping",    "unknown",
  "414",   "ICD8",   1,  "I25", "Chronic ischemic heart disease",                              "approx",
  #------------------------------------------------------
  # ICD7
  #------------------------------------------------------
  "444",   "ICD7",  28,  NA,    "Possibly respiratory condition, older code => check registry","unknown",
  "581",   "ICD7",  18,  "check local registry","Could be kidney/digestive code in older classification","unknown",
  "345",   "ICD7",  16,  "G40", "Epilepsy / seizures (older usage)",                           "approx",
  "201",   "ICD7",  12,  "C81", "Hodgkin lymphoma (older classification)",                    "approx",
  "224",   "ICD7",  10,  "check local registry","Older classification code => uncertain",     "unknown",
  "410",   "ICD7",  10,  "I21", "Acute myocardial infarction (older usage)",                  "approx",
  "411",   "ICD7",   8,  "I20", "Angina pectoris (older usage)",                              "approx"
)

icd_code_convert_key <- 
mapped_codes |> filter(certainty != "unknown") |> 
  bind_rows(convert_key)


write_csv(icd_code_convert_key, here::here("..", "data", "ICD_codes_convert_key_20250226.csv"))
```


- Extract ICD-diagnoses from patient registry and the death registry.
- Merge and select cases based on the base dataset.
- Create time based variables
  - From HPA one year forward
  - From HPA five year forward

```{r}

# icd convert key
convert_key <-
read_csv(
  here::here("..", "data", "ICD_codes_convert_key_20250226.csv")
)

# patient registers data , inpatient care
patreg_inpat <- haven::read_sas(here::here("..", "..", "..", "..",
"HPI (DNR XXXXXX)", "Data HPI registeruttag 2025",
"data_socialstyrelsen", "data_raw", "patientreg", "ut_r_par_sv_11104_2024.sas7bdat"))




```

	•	The code standardizes diagnosis codes by first converting all raw diagnosis entries (from multiple ICD versions) to a common format—that is, to the level of one letter followed by two digits (i.e. the ICD10 chapter level). any characters beyond that are dropped so that the analysis uses a consistent level of detail.
	•	It assigns the ICD version based on the admission year (AR):
• AR ≤ 1968 is assigned ICD7
• AR ≤ 1986 is assigned ICD8
• AR ≤ 1996 is assigned ICD9
• otherwise, ICD10 is assumed.
	•	A special rule is added for Skåne: because the region switched to ICD10 later (in 1998), if the admission year is 1997 and the diagnosis code does not start with a letter, it is recoded as ICD9.
	•	The code then “cleans” the diagnosis code by:
• For ICD7 and ICD8: keeping only the first three digits (discarding any comma and the following text).
• For ICD9: removing any trailing letter and additional characters (so that, for example, codes like “205B0”, “184A07”, or “410AA1” are reduced to just the first three digits).
	•	Finally, the cleaned dataset is merged with a conversion key (using a left join) that maps the original ICD codes to their approximate ICD10 counterparts.And pivoted back to wide. This final merged dataset (version_merged) enables consistent analysis across different periods by aligning older ICD codes to a modern ICD10 framework.
```{r}

patreg_inpat |> group_by(LopNr, HDIA, AR, INDATUMA,UTDATUMA) |> 
  summarise(n=n()) |> arrange(desc(n))

patreg_long <-
  patreg_inpat |> 
  # remove duplicates
  distinct() |> 
  mutate(
    indatum = ymd(INDATUMA),
    utdatum = ymd(UTDATUMA),
    #across(everything(), str_trim), 
    .after = AR,
    .keep = "unused",
    
  ) |> 
  pivot_longer(cols = contains("DIA"),
               names_to = "dia_version",
               values_to = "dia",
               values_drop_na = TRUE
               ) |> 
  filter(dia != "" & dia != "[BLANKAD]") |> 
  mutate(
    dia = str_trim(dia),
      icd_version = case_when(
      AR <= 1968 ~ "ICD7",
      AR <= 1986 ~ "ICD8",
      AR <= 1996 ~ "ICD9",
      TRUE ~ "ICD10" 
    ),
   icd_version = if_else(
# Skåne switched to icd 10 later than other, in 1998, so if there are icd codes that look like icd 9 in 1997
      AR == 1997 & !str_detect(dia, "^[A-Za-z]"), "ICD9", icd_version
    ),
# Converting codes to main and not subversions of the icd codes
      dia = case_when(
      # icd7: keep the first three digits, remove the comma and anything after
      icd_version == "ICD7" ~ str_replace(dia, "(\\d{3}),.*", "\\1"),
      # icd8: keep the first three digits, remove the comma and anything after
      icd_version == "ICD8" ~ str_replace(dia, "(\\d{3}),.*", "\\1"),
      # icd9: keep the three digits, remove any with one trailing letter, and any other trailing character after it
      icd_version == "ICD9" ~ str_replace(dia, "^(\\d{3})[A-Za-z].*", "\\1"),
       # ICD10: keep one letter followed by one or two digits, remove anything after
  icd_version == "ICD10" ~ str_replace(dia, "^([A-Z]\\d{1,2}).*", "\\1"),
        # if it's not icd7, icd8 or icd9, keep as is
      TRUE ~ dia
    )
  )

version_merged <-
  patreg_long |> 
  left_join(
    convert_key,
    by = c("dia" = "original_icd", "icd_version" = "converted_from_icd")
  ) |> 
  mutate(
      dia = coalesce(new_icd10_code, dia)
  ) |> 
  group_by(LopNr, AR, ALDER, indatum, utdatum, dia_version) %>%
  mutate(dia_id = row_number()) %>%
  ungroup() |> 
  # remove duplicates (can have been added after removing [BLANKAD] and recoding ICD codes)
  filter(dia_id == 1)

version_merged_wide <- version_merged |>
  select(LopNr, AR, ALDER, indatum, utdatum, dia, dia_version) |> 
  pivot_wider(
    id_cols = c(LopNr, AR, ALDER, indatum, utdatum),  
    names_from = c(dia_version),
    values_from = dia,
    values_fill = list(dia = NA)
  )




# not merged icd codes from icd 7,8,9
# version_merged |> filter(is.na(new_icd10_code) & !is.na(dia)) |> count(dia, AR, icd_version) |> 
#   filter(icd_version != "ICD10") |> arrange(desc(n)) |> 
#   group_by(dia, icd_version) |> 
#   summarise(n= sum(n))|> arrange(desc(icd_version),desc(n)) 

```



```{r}
 write_csv(version_merged_wide, here::here("..", "..", "..", "..",
"HPI (DNR XXXXXX)", "Data HPI registeruttag 2025",
"data_socialstyrelsen", "data_clean", "patientreg", "inpatient_data_clean_2025-02-26.csv"))



```



# death register

```{r}
# death register data
deathreg <- haven::read_sas(here::here("..", "..", "..", "..",
"HPI (DNR XXXXXX)", "Data HPI registeruttag 2025",
"data_socialstyrelsen", "data_raw", "dodsorsaksreg", "ut_r_dors_11104_2024.sas7bdat"))
```


```{r}
deathreg |> 
  pivot_longer(
    cols = contains("ORSAK"),
    names_to = "dia_version",
    values_to = "dia") |> 
  filter(dia != "") |> 
  filter(!str_detect(dia, "^[A-Za-z]\\d{2}$")) |> select(dia, ICD)


deathreg |>   filter(str_detect(DODSDAT, "00$")) |> select(DODSDAT)

```
Only main diagnoses are recoded
If both the month (positions 5–6) and the day (positions 7–8) are “00”, then recode month to “06” and day to “15”.
If only the month is “00”, then recode month to “06” while leaving the day unchanged.
If only the day is “00”, then recode day to “15” while leaving the month unchanged.

```{r}

deathreg_clean <-
deathreg |>
  mutate(ULORSAK = case_when(
    ULORSAK == "162" ~ "C34",  # malignant neoplasm of bronchus/lung
    ULORSAK == "173" ~ "C44",  # malignant neoplasm of skin (non-melanoma)
    ULORSAK == "174" ~ "C50",  # malignant neoplasm of the breast
    ULORSAK == "191" ~ "C71",  # malignant neoplasm of the brain
    ULORSAK == "250" ~ "E11",  # type 2 diabetes mellitus
    ULORSAK == "303" ~ "F10",  # alcohol-related disorders
    ULORSAK == "394" ~ "I05",  # rheumatic heart disease
    ULORSAK == "410" ~ "I21",  # acute myocardial infarction
    ULORSAK == "952" ~ "S06",  # intracranial injury
    ULORSAK == "_AV" ~ "Z99",  # administrative/other category
    TRUE ~ ULORSAK)) |> 
  select(LopNr, DODSDAT, ULORSAK, contains("MORSA")) |>
  mutate(
    DODSDAT = case_when(
      str_sub(DODSDAT,5,6)=="00" & str_sub(DODSDAT,7,8)=="00" ~ paste0(str_sub(DODSDAT,1,4), "06", "15"),
      str_sub(DODSDAT,5,6)=="00" ~ paste0(str_sub(DODSDAT,1,4), "06", str_sub(DODSDAT,7,8)),
      str_sub(DODSDAT,7,8)=="00" ~ paste0(str_sub(DODSDAT,1,6), "15"),
      TRUE ~ DODSDAT
    ),
    DODSDAT = ymd(DODSDAT)
  )



```






```{r}
 write_csv(deathreg_clean, here::here("..", "..", "..", "..",
"HPI (DNR XXXXXX)", "Data HPI registeruttag 2025",
"data_socialstyrelsen", "data_clean", "dodsorsaksreg", "deathreg_data_clean_2025-02-26.csv"))
```

