---
title: ""
format:
  html:
    embed-resources: true
    page-layout: full
    grid:
      margin-width: 100px
author: Daniel Väisänen
editor: source
toc: true
execute:
  warning: false
  echo: true
  eval: false
---


# Description
proportional‐hazards test for your sex, age interaction and occupation interaction models.  For each modifier you fit two nested models—one with a proportional treatment effect (anc = ~ treated) and one allowing the treatment spline to vary by that modifier (anc = ~ treated + treated:modifier)—then perform a likelihood‐ratio test.



```{r}
library(tidyverse)
library(flexsurv)
```





```{r}

# Model 1: proportional hazards by sex (no Sex in the spline ancillaries)
fit_sex_ph <- flexsurvspline(
  formula = Surv(risk_time_art_unrestricted, risk_art_flag_unrestricted) ~
    treated +
      Sex +
      Age +
      birth_cohort +
      comorbidity +
      cvd_art_before_HPA_flag +
      IncomeLevel_CSFVI +
      ssyk_WBHL_group,
  anc = list(
    gamma1 = ~treated,
    gamma2 = ~treated,
    gamma3 = ~treated
  ),
  data = data_cvd,
  k = 2,
  scale = "hazard"
)

# Model 2: allow treatment spline shape to differ by sex (non‐proportional hazards)
fit_sex_tvh <- flexsurvspline(
  formula = Surv(risk_time_art_unrestricted, risk_art_flag_unrestricted) ~
    treated +
      Sex +
      Age +
      birth_cohort +
      comorbidity +
      cvd_art_before_HPA_flag +
      IncomeLevel_CSFVI +
      ssyk_WBHL_group,
  anc = list(
    gamma1 = ~ treated + Sex,
    gamma2 = ~ treated + Sex,
    gamma3 = ~ treated + Sex
  ),
  data = data_cvd,
  k = 2,
  scale = "hazard"
)

# extract log‐likelihoods and number of parameters
ll_ph <- fit_sex_ph$loglik # –898492
ll_tvh <- fit_sex_tvh$loglik # –898434.8
npar_ph <- fit_sex_ph$npars # 28
npar_tvh <- fit_sex_tvh$npars # 31

# compute LRT statistic and degrees of freedom
lrt_stat <- 2 * (ll_tvh - ll_ph)
df_diff <- npar_tvh - npar_ph

# p‐value for chi‐square(df = 3)
p_value <- pchisq(lrt_stat, df = df_diff, lower.tail = FALSE)

# report
cat(
  "LRT χ² =",
  round(lrt_stat, 1),
  "on",
  df_diff,
  "df; p =",
  format.pval(p_value, digits = 3),
  "\n"
)
```

large χ² and tiny p‐value tells that allowing the spline shape of the treatment effect to differ by sex (i.e. non–proportional hazards) significantly improves model fit. In other words, the treatment hazard ratio does vary over time by sex, and we should retain the Sex terms in your anc = list(gamma…) if we wish to model that time‐varying behaviour.

```{r}
library(flexsurv)

# Model A1: proportional hazards by age (no Age in the spline ancillaries)
fit_age_ph <- flexsurvspline(
  formula = Surv(risk_time_art_unrestricted, risk_art_flag_unrestricted) ~
    treated +
      Age +
      Sex +
      birth_cohort +
      comorbidity +
      cvd_art_before_HPA_flag +
      IncomeLevel_CSFVI +
      ssyk_WBHL_group,
  anc = list(
    gamma1 = ~treated,
    gamma2 = ~treated,
    gamma3 = ~treated
  ),
  data = data_cvd,
  k = 2,
  scale = "hazard"
)

# Model A2: allow treatment spline shape to differ by age (non‐PH by age)
fit_age_tvh <- flexsurvspline(
  formula = Surv(risk_time_art_unrestricted, risk_art_flag_unrestricted) ~
    treated +
      Age +
      Sex +
      birth_cohort +
      comorbidity +
      cvd_art_before_HPA_flag +
      IncomeLevel_CSFVI +
      ssyk_WBHL_group,
  anc = list(
    gamma1 = ~ treated + treated:Age,
    gamma2 = ~ treated + treated:Age,
    gamma3 = ~ treated + treated:Age
  ),
  data = data_cvd,
  k = 2,
  scale = "hazard"
)

# LRT for age
ll_ph_age <- fit_age_ph$loglik
ll_tvh_age <- fit_age_tvh$loglik
npar_ph_age <- fit_age_ph$npars
npar_tvh_age <- fit_age_tvh$npars

lrt_age <- 2 * (ll_tvh_age - ll_ph_age)
df_age <- npar_tvh_age - npar_ph_age
p_age <- pchisq(lrt_age, df = df_age, lower.tail = FALSE)

cat(
  "Age LRT χ² =",
  round(lrt_age, 1),
  "on",
  df_age,
  "df; p =",
  format.pval(p_age, 3),
  "\n"
)
```



```{r}
# Model O1: proportional hazards by occupation (no occupation in ancillaries)
fit_occ_ph <- flexsurvspline(
  formula = Surv(risk_time_art_unrestricted, risk_art_flag_unrestricted) ~
    treated +
      ssyk_WBHL_group +
      Sex +
      Age +
      birth_cohort +
      comorbidity +
      cvd_art_before_HPA_flag +
      IncomeLevel_CSFVI,
  anc = list(
    gamma1 = ~treated,
    gamma2 = ~treated,
    gamma3 = ~treated
  ),
  data = data_cvd,
  k = 2,
  scale = "hazard"
)

# Model O2: allow treatment spline shape to differ by occupation (non‐PH by occupation)
fit_occ_tvh <- flexsurvspline(
  formula = Surv(risk_time_art_unrestricted, risk_art_flag_unrestricted) ~
    treated +
      ssyk_WBHL_group +
      Sex +
      Age +
      birth_cohort +
      comorbidity +
      cvd_art_before_HPA_flag +
      IncomeLevel_CSFVI,
  anc = list(
    gamma1 = ~ treated + treated:ssyk_WBHL_group,
    gamma2 = ~ treated + treated:ssyk_WBHL_group,
    gamma3 = ~ treated + treated:ssyk_WBHL_group
  ),
  data = data_cvd,
  k = 2,
  scale = "hazard"
)

# LRT for occupation
ll_ph_occ <- fit_occ_ph$loglik
ll_tvh_occ <- fit_occ_tvh$loglik
npar_ph_occ <- fit_occ_ph$npars
npar_tvh_occ <- fit_occ_tvh$npars

lrt_occ <- 2 * (ll_tvh_occ - ll_ph_occ)
df_occ <- npar_tvh_occ - npar_ph_occ
p_occ <- pchisq(lrt_occ, df = df_occ, lower.tail = FALSE)

cat(
  "Occupation LRT χ² =",
  round(lrt_occ, 1),
  "on",
  df_occ,
  "df; p =",
  format.pval(p_occ, 3),
  "\n"
)
```
