---
title: ""
format:
  html:
    embed-resources: true
    page-layout: full
    grid:
      margin-width: 100px
author: Daniel Väisänen
editor: source
toc: true
execute:
  warning: false
  echo: true
  eval: false
---

# Description
Cox-models output for figure 1.


# Libraries
```{r}
library(flexsurv)
library(tidyverse)
library(marginaleffects)
library(survival)
library(broom)

```


# data

```{r}
source(here::here("script", "01_data_source_file.R"))
source(here::here("script", "02_functions_source_file.R"))



# helper function to get HRs and counts -------------------------------
get_hr <- function(fit, by_var) {

  ## 1 ── hazard ratio (treated vs control) ----------------------------
  hr <- fit |>
    avg_comparisons(
      variables = "treated",
      by        = by_var,
      type      = "lp",
      transform = "exp")

  ## 2 ── arm-specific counts -----------------------------------------
  mf       <- model.frame(fit)          # data used by coxph
  surv_obj <- mf[[1]]                   # the Surv() object (time, status)

  counts <- mf |>
    mutate(event = surv_obj[, "status"]) |>                 # 1 = event
    mutate(arm   = if_else(treated %in% c(1, "HPA"),        # label arms
                           "HPA", "Control")) |>
    group_by(.data[[by_var]], arm) |>
    summarise(n       = n(),
              events  = sum(event),
              .groups = "drop") |>
    tidyr::pivot_wider(                                     # → n_HPA, n_Control …
      names_from  = arm,
      values_from = c(n, events),
      names_glue  = "{.value}_{arm}"        # n_HPA, events_HPA, n_Control, ...
    )

  ## 3 ── merge --------------------------------------------------------
  dplyr::left_join(hr, counts, by = by_var)
}

```

# ───────────────────────────────────────────────────────────────────────────────

# Cox main models and sociodemographics

## data
```{r}
data_occ  <- data_cvd |>
  filter(!is.na(ssyk_WBHL_group)) |>
  group_by(id_cluster) |>
        mutate(n_cluster = n()) |>
        ungroup() |>
  filter(n_cluster > 4)

data_mun  <- data_cvd |>
  filter(!is.na(ssyk_WBHL_group),
         !is.na(KommunSize),
         !is.na(Place_of_origin)) |>
  group_by(id_cluster) |>
        mutate(n_cluster = n()) |>
        ungroup() |>
  filter(n_cluster > 4)


```

## models

We fit the models grouped by their role rather than as a flat, positionally
indexed list. This makes the single pre-specified primary model explicit and
prevents the earlier bug where a model meant one thing here and was extracted
as another downstream. Bias-probe / lifestyle interactions (HPA-only) live in
their own script (26b) so they cannot leak into the main results.

```{r}
# fit_primary: the single pre-specified primary model (M3), built from the
# one canonical M3 formula defined in 02_functions_source_file.R so it can
# never drift from the M3 fitted in 35_ or the competing-risks script.
fit_primary <- coxph(m3_formula_10yr, data = data_cvd)

# fits_sensitivity: adjustment-set sensitivity models around the primary.
#   m1 = birth cohort only; m2 = + age/sex/comorbidity/prior CVD;
#   m4 = + occupation; m5 = + municipality & origin.
fits_sensitivity <- list(
  m1 = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated + birth_cohort,
    data = data_cvd
  ),
  m2 = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated + birth_cohort + Age + Sex +
      comorbidity + cvd_art_before_HPA_flag,
    data = data_cvd
  ),
  m4 = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated + ssyk_WBHL_group + birth_cohort + Age + Sex +
      comorbidity + cvd_art_before_HPA_flag + IncomeLevel_CSFVI,
    data = data_occ
  ),
  m5 = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated + birth_cohort + Age + Sex +
      comorbidity + cvd_art_before_HPA_flag + IncomeLevel_CSFVI +
      ssyk_WBHL_group + KommunSize + Place_of_origin,
    data = data_mun
  )
)

# fits_effectmod: PRE-SPECIFIED effect-modification questions (sex, age,
# education), each a treated * modifier interaction on the M3 adjustment set.
fits_effectmod <- list(
  sex = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated * Sex + birth_cohort + Age +
      comorbidity + cvd_art_before_HPA_flag + IncomeLevel_CSFVI,
    data = data_cvd
  ),
  age = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated * age_group + Sex + birth_cohort +
      comorbidity + cvd_art_before_HPA_flag + IncomeLevel_CSFVI,
    data = data_cvd
  ),
  education = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated * EducationLevel + birth_cohort + Age + Sex +
      comorbidity + cvd_art_before_HPA_flag,
    data = data_cvd
  )
)

# fits_effectmod_exploratory: additional, NOT pre-specified effect-modification
# (income, detailed occupation) kept for completeness but labelled exploratory.
fits_effectmod_exploratory <- list(
  income = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated * IncomeLevel_CSFVI + birth_cohort + Age + Sex +
      comorbidity + cvd_art_before_HPA_flag,
    data = data_cvd
  ),
  occupation = coxph(
    Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~
      treated * SSYK + birth_cohort + Age + Sex +
      comorbidity + cvd_art_before_HPA_flag,
    data = data_occ |>
      filter(SSYK != "Military") |>
      mutate(SSYK = fct_drop(SSYK))
  )
)

# Save one named container; downstream scripts read elements by name, never
# by position.
cox_ten_year_models <- list(
  fit_primary                 = fit_primary,
  fits_sensitivity            = fits_sensitivity,
  fits_effectmod              = fits_effectmod,
  fits_effectmod_exploratory  = fits_effectmod_exploratory
)

write_rds(
  cox_ten_year_models,
  here::here("..", "results", "models", "cox_ten_year_models.rds")
)
```

## main model (M1–M5) estimates

```{r}
# helper just for the mains
get_main <- function(fit, model_id) {
  td <- tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "treatedHPA") %>%
    select(term, estimate, std.error, statistic, p.value, starts_with("conf")) %>%
    mutate(model = model_id, .before = 1)

  mf     <- model.frame(fit)
  status <- mf[[1]][, "status"]
  dat    <- mf %>%
    mutate(event = status,
           arm   = if_else(treated %in% c(1, "HPA"),
                           "HPA", "Control"))

  counts <- dat %>%
    group_by(arm) %>%
    summarise(n      = n(),
              events = sum(event),
              .groups = "drop") %>%
    pivot_wider(names_from  = arm,
                values_from = c(n, events),
                names_glue  = "{.value}_{arm}")

  bind_cols(td, counts)
}

# Assemble the five main models in display order. M3 is the primary fit; the
# others come from fits_sensitivity. The m1..m5 codes are kept only as the
# stable identifiers the forest-plot script (32) expects.
fits_main <- list(
  m1 = fits_sensitivity$m1,
  m2 = fits_sensitivity$m2,
  m3 = fit_primary,
  m4 = fits_sensitivity$m4,
  m5 = fits_sensitivity$m5
)

tbl_main <- imap_dfr(fits_main, \(fit, nm) get_main(fit, nm))
```



## effect-modification estimates
```{r}
# Map each named effect-modification model to (a) the stratifying variable it
# is averaged over and (b) the stable exposure code used by the forest plot.
# sex/age/education are pre-specified; income/occupation are exploratory.
effectmod_fits <- c(fits_effectmod, fits_effectmod_exploratory)

effectmod_by <- c(
  sex        = "Sex",
  age        = "age_group",
  education  = "EducationLevel",
  income     = "IncomeLevel_CSFVI",
  occupation = "SSYK"
)

effectmod_code <- c(
  sex        = "m6",
  age        = "m7",
  education  = "m10",
  income     = "m9",
  occupation = "m11"
)

tbl_int_sociodemographics <- imap_dfr(
  effectmod_fits,
  \(fit, nm) {
    by_var <- effectmod_by[[nm]]
    get_hr(fit, by_var) |>
      mutate(exposure = effectmod_code[[nm]], .before = 1)
  }
)
```


# ───────────────────────────────────────────────────────────────────────────────
# Bias-probe / lifestyle interactions live in 26b_bias_probes.qmd
# (HPA-only variables, controls forced to the reference category). They are kept
# out of this main-results pipeline on purpose so they cannot leak into the
# pre-specified results.
# ───────────────────────────────────────────────────────────────────────────────


# join main results (for the forest plot)

```{r}
tbl_main
tbl_int_sociodemographics |> as_tibble()


# 1. MAIN MODELS (M1–M5) ------------------------------------------------
tbl_main2 <- tbl_main |>
  rename(exposure = model) |>          # rename "model" → "exposure"
  mutate(groups = exposure) |>         # one overall stratum
  select(
    exposure, groups,
    n_Control, events_Control,
    n_HPA, events_HPA,
    term,
    estimate, conf.low, conf.high, p.value
  )

# 2. EFFECT MODIFICATION (sex/age/education pre-specified; income/occupation
#    exploratory) — already carries the m6/m7/m9/m10/m11 codes the plot uses.
tbl_int2 <- tbl_int_sociodemographics |>
  mutate(
    groups = coalesce(                 # pick whichever stratum is non-NA
      Sex, age_group, SSYK, EducationLevel, IncomeLevel_CSFVI
    )
  ) |>
  select(
    exposure, groups,
    n_Control, events_Control,
    n_HPA, events_HPA,
    term, contrast,
    estimate, conf.low, conf.high, p.value
  )

# 3. BIND main + effect modification -----------------------------------
final_hr_tbl <- bind_rows(tbl_main2, tbl_int2)

final_main <- final_hr_tbl %>%
  mutate(
    block = case_when(
      exposure %in% paste0("m", 1:5) ~ "Main models",
      exposure %in% c("m6", "m7", "m9", "m10", "m11") ~ "Sociodemographics",
      TRUE ~ "Other"
    ),
    label = case_when(
      exposure == "m1" ~ "M1: birth cohort",
      exposure == "m2" ~ "M2: + age, sex, comorbidity & CVD",
      exposure == "m3" ~ "M3: + income",
      exposure == "m4" ~ "M4: + occupation",
      exposure == "m5" ~ "M5: + municipality & origin",
      exposure == "m7" ~ paste0(groups, " yrs"),
      exposure %in% c("m6", "m9", "m10", "m11") ~ paste0(groups),
      TRUE ~ groups
    )
  ) %>%
  mutate(
    block = factor(
      block,
      levels = c("Main models", "Sociodemographics")
    )
  ) %>%
  arrange(block, exposure, groups) %>%
  mutate(row = row_number())


write_csv(
  final_main,
  here::here("..", "results", "models", "cox_ten_year_main_HR.csv")
)
```


# ───────────────────────────────────────────────────────────────────────────────


# example of relative and absolute risks

```{r}

# coxph(Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~ treated, data = data_cvd |>   mutate(
#     n_tests_within_5_fac_vs_ctrl = fct_drop(n_tests_within_5_fac_vs_ctrl)) |>
#       filter(n_tests_within_5_fac_vs_ctrl == "HPA, 1 test")
#   ) |> tidy(exponentiate = TRUE)
#
# coxph(Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~ treated, data = data_cvd |>   mutate(
#     n_tests_within_5_fac_vs_ctrl = fct_drop(n_tests_within_5_fac_vs_ctrl)) |>
#       filter(n_tests_within_5_fac_vs_ctrl == "HPA, 2 tests")
#   ) |> tidy(exponentiate = TRUE)
#
# coxph(Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~ treated, data = data_cvd |>   mutate(
#     n_tests_within_5_fac_vs_ctrl = fct_drop(n_tests_within_5_fac_vs_ctrl)) |>
#       filter(n_tests_within_5_fac_vs_ctrl == "HPA, ≥3 tests")
#   ) |> tidy(exponentiate = TRUE)

# Model
co_fit <-
coxph(Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~ n_tests_within_5_fac_vs_ctrl*treated +
        BMI_cat + Sex +
    Age + birth_cohort + comorbidity + cvd_art_before_HPA_flag +
    IncomeLevel_CSFVI, data = data_cvd |>   mutate(
    n_tests_within_5_fac_vs_ctrl = fct_drop(n_tests_within_5_fac_vs_ctrl)
  ))

# relative effect (hazard ratio)
# average hazard ratios (treated vs control) within each test stratum
# “in the ≥3 test stratum, hpa is associated with a 29 % lower hazard of art within 10 years (hr 0.71, 95 % ci 0.65–0.77).”
# within the stratum, again with individual covariates fixed


hr_tbl <- co_fit |>
  avg_comparisons(                 # average contrasts
    variables = "treated",         # focal variable
    by        = "n_tests_within_5_fac_vs_ctrl",
    type      = "lp",              # linear predictor = log-hazard
    transform = "exp")             # exponentiate → hazard ratio

hr_tbl


# predicted 10-year risk within stratum
# scale: probability
# “among individuals classified as ‘hpa, ≥3 tests’, the marginal predicted 10-year risk is 0.77 (95 % ci 0.74–0.80).”
#  averaged over their observed distributions of bmi, sex, age, etc.

pre_tbl <- avg_predictions(
  co_fit,
  variables = "treated",
  by = "n_tests_within_5_fac_vs_ctrl",
  type = "risk")
pre_tbl

# absolute effect (risk difference)
# scale: probability
# “moving a person from control to hpa status, while keeping the test stratum fixed at ≥3, reduces the 10-year risk by 0.27 (95 % ci 0.22–0.31).”
#  while holding their own bmi, sex, age, etc. fixed and keeping them in the same test stratum
abs_tbl <- avg_comparisons(
  co_fit,
  variables = "treated",
  by = "n_tests_within_5_fac_vs_ctrl",
  type = "risk")
abs_tbl



```
