---
title: ""
format:
  html:
    embed-resources: true
    page-layout: full
    grid:
      margin-width: 100px
author: Daniel Väisänen
editor: source
toc: true
execute:
  warning: false
  echo: true
  eval: false
---

# Description

Exploratory **bias probes**, separated from the main results pipeline (26_).

These lifestyle variables (VO2max, BMI, exercise, self-rated health) and the
number-of-assessments dose-response are measured **only in HPA participants**;
controls inherit the matched HPA participant's category within the cluster. They
therefore cannot estimate a causal effect — they are reported, per the reviewers,
as exploratory probes of whether the association tracks plausible bias / selection
gradients, not as effect-modification or dose-response evidence. Keeping them in
their own script and their own output file (`ten_yr_biasprobe_interaction_HR.csv`)
prevents them leaking into the pre-specified results.


# Libraries
```{r}
library(flexsurv)
library(tidyverse)
library(marginaleffects)
library(survival)
```


# Data
```{r}
source(here::here("script", "01_data_source_file.R"))
source(here::here("script", "02_functions_source_file.R"))


# helper to get HRs and arm-specific counts (same as 26_) ---------------
get_hr <- function(fit, by_var) {
  hr <- fit |>
    avg_comparisons(
      variables = "treated",
      by        = by_var,
      type      = "lp",
      transform = "exp"
    )

  mf       <- model.frame(fit)
  surv_obj <- mf[[1]]

  counts <- mf |>
    mutate(event = surv_obj[, "status"]) |>
    mutate(arm = if_else(treated %in% c(1, "HPA"), "HPA", "Control")) |>
    group_by(.data[[by_var]], arm) |>
    summarise(n = n(), events = sum(event), .groups = "drop") |>
    tidyr::pivot_wider(
      names_from  = arm,
      values_from = c(n, events),
      names_glue  = "{.value}_{arm}"
    )

  dplyr::left_join(hr, counts, by = by_var)
}
```


# Fit one interaction model per probe
```{r}
# exposure name -> the HPA-only variable it interacts with treated
exposures <- c(
  vo2    = "vo2_cat",
  bmi    = "BMI_cat",
  exer   = "Exercise_cat",
  health = "Health_cat",
  tests  = "n_tests_within_5_fac_vs_ctrl"
)

# fixed adjustment set (the M3 covariates minus treated, which enters via the
# interaction term)
cov_base <- paste(
  setdiff(m3_covariates, "treated"),
  collapse = " + "
)

# fit one Cox model per probe, each with a treated * probe interaction.
fits_biasprobe <- imap(
  exposures,
  \(var, nm) {
    fml <- as.formula(
      glue::glue(
        "Surv(risk_time_art_10yrs, risk_art_flag_10yrs) ~ treated * {var} + {cov_base}"
      )
    )

    # drop rows missing the probe itself, then keep clusters with >= 5 usable rows
    clean_dat <- data_cvd |>
      drop_na(all_of(var)) |>
      group_by(id_cluster) |>
      filter(n() >= 5) |>
      ungroup()

    coxph(fml, data = clean_dat)
  }
)

# write the fitted objects AFTER the loop completes (previously this sat inside
# the loop and wrote a half-built object).
write_rds(
  fits_biasprobe,
  here::here("..", "results", "models", "ten_yr_biasprobe_interaction_fits.rds")
)
```


# Extract HRs into one tidy table
```{r}
hr_tbl_all <- imap_dfr(
  fits_biasprobe,
  \(fit, nm) {
    by_var <- exposures[[nm]]
    get_hr(fit, by_var) |>
      mutate(exposure = nm, .before = 1)
  }
)

# build the forest-plot-ready rows: stable exposure codes, block, label.
biasprobe_hr <- hr_tbl_all |>
  as_tibble() |>
  mutate(
    groups = coalesce(
      vo2_cat, BMI_cat, Exercise_cat, Health_cat, n_tests_within_5_fac_vs_ctrl
    ),
    exposure = recode(
      exposure,
      vo2    = "VO2",
      bmi    = "BMI",
      exer   = "Exercise",
      health = "Health",
      tests  = "Tests"
    ),
    block = if_else(exposure == "Tests", "Number of tests", "Lifestyle related"),
    label = case_when(
      exposure == "Tests"    ~ paste0(groups),
      exposure == "VO2"      ~ paste0("VO2max: ", groups),
      exposure == "BMI"      ~ paste0("BMI: ", groups),
      exposure == "Exercise" ~ paste0("Exercise: ", groups),
      exposure == "Health"   ~ paste0("Health: ", groups),
      TRUE                   ~ groups
    )
  ) |>
  select(
    exposure, block, label, groups,
    n_Control, events_Control, n_HPA, events_HPA,
    term, contrast, estimate, conf.low, conf.high, p.value
  )

biasprobe_hr |> print(n = 200)

write_csv(
  biasprobe_hr,
  here::here("..", "results", "models", "ten_yr_biasprobe_interaction_HR.csv")
)
```
