---
title: ""
format:
  html:
    embed-resources: true
    page-layout: full
    grid:
      margin-width: 100px
author: Daniel Väisänen
editor: source
toc: true
execute:
  warning: false
  echo: true
  eval: false
---

# Description

Competing-risk sensitivity analysis (referees 2.8, 2.12). The pre-specified
primary analysis is the **cause-specific** Cox model that censors at death.
Here we additionally treat death as a competing event and report the
**subdistribution hazard ratio** (Fine–Gray) plus cumulative incidence functions
that account for death. This is a sensitivity analysis, not a replacement for the
primary cause-specific estimate.

The covariate set is the single canonical M3 set defined in
`02_functions_source_file.R`.


# Libraries
```{r}
library(tidyverse)
library(survival)   # finegray(), coxph(), multistate survfit()
library(broom)
```


# Data
```{r}
source(here::here("script", "01_data_source_file.R"))
source(here::here("script", "02_functions_source_file.R"))
```


# Build the competing-event status
```{r}
# Within the 10-year window:
#   1 = CVD event   (the cause-specific event flag)
#   2 = death       (no CVD, and follow-up ended because of death)
#   0 = censored    (administrative / end of follow-up)
#
# risk_end_art_10yrs is pmin(CVD date, death date, admin censor); so when there
# is no CVD and death == risk_end, death is the terminal event. If the admin
# censor came first, DODSDAT > risk_end and the person is censored.
data_cr <- data_cvd |>
  mutate(
    DODSDAT            = as.Date(DODSDAT),
    risk_end_art_10yrs = as.Date(risk_end_art_10yrs),
    cr_status = case_when(
      risk_art_flag_10yrs == 1 ~ 1L,
      !is.na(DODSDAT) & DODSDAT <= risk_end_art_10yrs ~ 2L,
      TRUE ~ 0L
    ),
    cr_status = factor(cr_status, levels = c(0, 1, 2),
                       labels = c("censored", "cvd", "death"))
  )

# verification: the three categories must partition the cohort
data_cr |> count(treated, cr_status)
data_cr |>
  summarise(n = n(), sum_status = sum(table(cr_status))) # sum must equal n
```


# Subdistribution hazard ratio (Fine–Gray) — memory-safe

`survival::finegray()` physically expands the data to subdistribution risk sets,
which on ~2 million rows can exhaust memory. We therefore prefer
**`fastcmprsk::fastCrr()`**, which fits the Fine–Gray model with an O(n)
forward–backward scan and **no row expansion**, so it runs on the full cohort with
modest memory. If `fastcmprsk` is not installed, the code falls back to fitting the
model on a **random subsample of whole matched clusters** (preserving the 1:4
matching and the estimand) via `survival::finegray()`.

`fastcmprsk` and its dependency `dynpred` are CRAN-archived; install from the CRAN
archive with:
`remotes::install_version("dynpred"); remotes::install_version("fastcmprsk")`.

```{r}
# prefer the full-data O(n) Fine–Gray; fall back to a memory-safe subsample.
use_fastcmprsk <- requireNamespace("fastcmprsk", quietly = TRUE)

if (use_fastcmprsk) {
  library(fastcmprsk)

  # numeric status for fastcmprsk: 0 = censored, 1 = CVD, 2 = death
  fc_df <- data_cr |>
    mutate(fstatus = as.integer(cr_status) - 1L) |>
    select(all_of(c("risk_time_art_10yrs", "fstatus", m3_covariates))) |>
    drop_na()

  message(sprintf("[27b] fastcmprsk Fine-Gray on full data (%s subjects).",
                  scales::comma(nrow(fc_df))))

  fc_form <- as.formula(
    paste("Crisk(risk_time_art_10yrs, fstatus, cencode = 0, failcode = 1) ~", m3_rhs)
  )

  # Variance is bootstrap-based. useMultipleCores = TRUE forks the process and
  # each fork works on the full ~2M-row data, which can exhaust RAM and crash R —
  # so we keep it single-core. B is the number of bootstrap reps: lower it (e.g.
  # 50) if the CI step is too slow, raise it for a more stable CI.
  fc_fit <- fastCrr(
    fc_form,
    data        = fc_df,
    variance    = TRUE,
    var.control = varianceControl(B = 100, useMultipleCores = FALSE)
  )

  # fastCrr returns UNNAMED coefficients in model-matrix column order; label them
  # so we can pull the treated term reliably.
  cov_names <- colnames(model.matrix(as.formula(paste("~", m3_rhs)), data = fc_df))[-1]
  b  <- setNames(fc_fit$coef, cov_names)
  se <- setNames(sqrt(diag(fc_fit$var)), cov_names)
  i  <- which(names(b) == "treatedHPA")

  shr_tbl <- tibble(
    analysis  = "Fine-Gray subdistribution HR (death competing; fastcmprsk, full data)",
    estimate  = exp(b[i]),
    conf.low  = exp(b[i] - 1.96 * se[i]),
    conf.high = exp(b[i] + 1.96 * se[i]),
    p.value   = 2 * pnorm(-abs(b[i] / se[i]))
  )
} else {
  # fallback: subsample whole clusters, then survival::finegray() + weighted Cox
  set.seed(20240601)
  target_clusters <- 60000  # ~300,000 subjects at 1:4; raise if memory allows

  keep_clusters <- data_cr |>
    distinct(id_cluster) |>
    slice_sample(n = min(target_clusters, dplyr::n_distinct(data_cr$id_cluster))) |>
    pull(id_cluster)

  fg_input <- data_cr |>
    filter(id_cluster %in% keep_clusters) |>
    select(all_of(c("cr_status", "risk_time_art_10yrs", m3_covariates))) |>
    drop_na()  # NB: id_cluster intentionally excluded so it is not a covariate

  message(sprintf(
    "[27b] fastcmprsk not found — finegray on %s subjects from %s clusters (subsample).",
    scales::comma(nrow(fg_input)), scales::comma(length(keep_clusters))
  ))

  fg_cvd <- finegray(
    Surv(risk_time_art_10yrs, cr_status) ~ .,
    data  = fg_input,
    etype = "cvd"
  )

  fit_fg <- coxph(
    build_m3_formula("Surv(fgstart, fgstop, fgstatus)"),
    data    = fg_cvd,
    weights = fgwt
  )

  shr_tbl <- broom::tidy(fit_fg, exponentiate = TRUE, conf.int = TRUE) |>
    filter(term == "treatedHPA") |>
    transmute(
      analysis = sprintf(
        "Fine-Gray subdistribution HR (death competing; %s-cluster subsample)",
        scales::comma(length(keep_clusters))
      ),
      estimate, conf.low, conf.high, p.value
    )
}

shr_tbl

write_csv(
  shr_tbl,
  here::here("..", "results", "tables", "competing_risks_shr.csv")
)
```


# Cumulative incidence functions accounting for death
```{r}
# Aalen–Johansen cumulative incidence by arm (survival >= 3.x multistate).
cif_fit <- survfit(Surv(risk_time_art_10yrs, cr_status) ~ treated, data = data_cr)

# tidy the per-state cumulative incidence (probability in each competing state)
cif_df <- broom::tidy(cif_fit) |>
  filter(state %in% c("cvd", "death")) |>
  mutate(
    years = time / 365.25,
    arm   = str_remove(strata, "treated="),
    state = recode(state, cvd = "CVD", death = "Death")
  )

cif_plot <- cif_df |>
  ggplot(aes(x = years, y = estimate, colour = arm)) +
  geom_step() +
  facet_wrap(~state, scales = "free_y") +
  labs(
    x = "Years since first HPA",
    y = "Cumulative incidence",
    colour = NULL,
    title = "Cumulative incidence of CVD and death by arm"
  ) +
  theme_minimal()

cif_plot

ggsave(
  here::here("..", "results", "figures", "competing_risks_cif.pdf"),
  plot = cif_plot, width = 210, height = 120, units = "mm", dpi = 300
)
```
