---
title: "Main-model marginal effects and effect-modification contrasts"
format: html
---

# Description

Population-averaged (marginal) hazard ratios for the main models and the
effect-modification contrasts. Models are read from the **named** container
written by 26_ and are addressed by name (e.g. `fits_effectmod$education`),
never by position — this removes the earlier bug where `m10` was Education here
but was extracted as the dose-response model.


# Libraries
```{r}
library(flexsurv)
library(tidyverse)
library(marginaleffects)
library(survival)
```


# Data
```{r}
source(here::here("script", "01_data_source_file.R"))
source(here::here("script", "02_functions_source_file.R"))
```


# Models
```{r}
models <- read_rds(
  here::here("..", "results", "models", "cox_ten_year_models.rds")
)

# named main models in display order (M3 is the primary fit)
fits_main <- list(
  m1 = models$fits_sensitivity$m1,
  m2 = models$fits_sensitivity$m2,
  m3 = models$fit_primary,
  m4 = models$fits_sensitivity$m4,
  m5 = models$fits_sensitivity$m5
)

# all effect-modification models, addressed by name
fits_effectmod <- c(models$fits_effectmod, models$fits_effectmod_exploratory)

# the stratifying variable for each effect-modification model
effectmod_by <- c(
  sex        = "Sex",
  age        = "age_group",
  education  = "EducationLevel",
  income     = "IncomeLevel_CSFVI",
  occupation = "SSYK"
)

# marginaleffects cannot recover the fitting data from a saved coxph object when
# called inside a loop, so we recreate each model's data and pass it via newdata.
# These subsets must match those used in 26_ (m4/m5 on occupation/municipality;
# the occupation interaction drops the Military group).
data_occ <- data_cvd |>
  filter(!is.na(ssyk_WBHL_group)) |>
  group_by(id_cluster) |>
  mutate(n_cluster = n()) |>
  ungroup() |>
  filter(n_cluster > 4)

data_mun <- data_cvd |>
  filter(!is.na(ssyk_WBHL_group), !is.na(KommunSize), !is.na(Place_of_origin)) |>
  group_by(id_cluster) |>
  mutate(n_cluster = n()) |>
  ungroup() |>
  filter(n_cluster > 4)

data_occ_mil <- data_occ |>
  filter(SSYK != "Military") |>
  mutate(SSYK = fct_drop(SSYK))

# data each model was fitted on (used as newdata for marginal effects)
main_data <- list(
  m1 = data_cvd, m2 = data_cvd, m3 = data_cvd, m4 = data_occ, m5 = data_mun
)
effectmod_data <- list(
  sex = data_cvd, age = data_cvd, education = data_cvd,
  income = data_cvd, occupation = data_occ_mil
)
```


# Marginal HR for the main models (M1–M5)
```{r}
main_effects <- imap_dfr(
  fits_main,
  \(fit, nm) {
    fit |>
      avg_comparisons(
        variables = "treated",
        newdata   = main_data[[nm]],
        type      = "lp",
        transform = "exp"
      ) |>
      as_tibble() |>
      transmute(
        model = nm,
        term, contrast,
        estimate, conf.low, conf.high, p.value
      )
  }
)

main_effects |>
  mutate(hr_ci = sprintf("%.2f (%.2f-%.2f)", estimate, conf.low, conf.high))
```


# Stratum-specific marginal HR for the effect-modification models
```{r}
effectmod_strata <- imap_dfr(
  fits_effectmod,
  \(fit, nm) {
    by_var <- effectmod_by[[nm]]
    fit |>
      avg_comparisons(
        variables = "treated",
        by        = by_var,
        newdata   = effectmod_data[[nm]],
        type      = "lp",
        transform = "exp"
      ) |>
      as_tibble() |>
      mutate(model = nm, stratifier = by_var, .before = 1)
  }
)

effectmod_strata |> print(n = 100)

# one tidy object holding both pieces
results <- list(
  main_effects     = main_effects,
  effectmod_strata = effectmod_strata
)

write_rds(
  results,
  here::here("..", "results", "models", "cox_ten_year_marginal_effects.rds")
)
```


# Pairwise interaction contrasts (ratio of HRs across modifier levels)
```{r}
get_interaction_comparisons <- function(fit, var_name, data) {
  # get levels from the supplied data (not model.frame, which can't recover the
  # data from a saved coxph object)
  v <- data[[var_name]]
  levs <- if (is.factor(v)) levels(droplevels(v)) else sort(unique(v))

  combos <- combn(levs, 2, simplify = FALSE)

  purrr::map_dfr(
    combos,
    \(pair) {
      vars <- setNames(list(c(pair[1], pair[2])), var_name)

      fit |>
        avg_comparisons(
          variables  = vars,
          by         = "treated",
          newdata    = data,
          comparison = "lnoravg",  # average log ratio of HRs
          transform  = "exp"
        ) |>
        as_tibble() |>
        mutate(
          interaction = var_name,
          comparison  = paste(pair[1], "vs", pair[2])
        )
    }
  )
}

# Pairwise contrasts only for the PRE-SPECIFIED modifiers (sex, age, education).
# Income/occupation are excluded: occupation has too many levels for meaningful
# pairwise contrasts and, being fit on a piped data subset, marginaleffects
# cannot recover its data for the counterfactual form used here.
all_interactions <- imap_dfr(
  models$fits_effectmod,
  \(fit, nm) {
    get_interaction_comparisons(fit, effectmod_by[[nm]], effectmod_data[[nm]]) |>
      mutate(model = nm, .before = 1)
  }
)

all_interactions |> print(n = 100)

write_csv(
  all_interactions,
  here::here(
    "..",
    "results",
    "tables",
    "cox_ten_year_interactions_main_estimates.csv"
  )
)
```


# Formatted interaction table
```{r}
table_df <- all_interactions |>
  mutate(
    estimate  = sprintf("%.2f", estimate),
    conf.low  = sprintf("%.2f", conf.low),
    conf.high = sprintf("%.2f", conf.high),
    or_ci     = paste0(estimate, " (", conf.low, "–", conf.high, ")")
  ) |>
  select(model, interaction, comparison, treated, or_ci) |>
  pivot_wider(
    names_from  = treated,
    values_from = or_ci
  ) |>
  arrange(model, interaction, comparison)

print(table_df, n = nrow(table_df))
```
