---
title: "Instantenous 10 year risk of CVD in HPA participants vs controls"
execute:
  warning: false
  echo: true
  eval: true
---

# Description

Forrestplot style plot of all different cox-models.

# Load libraries
```{r}
library(tidyverse)
library(ckbplotr)
library(glue)
library(scales)
library(showtext)
library(systemfonts)
```

# Setup fonts
```{r}
# We deliberately do NOT use showtext + Google Fonts here. ckbplotr renders the
# row labels with marquee, which uses systemfonts to find a BOLD face for
# **heading** markdown; showtext registers a family that marquee can't see a bold
# face for, so headings wouldn't render bold (and Google Fonts needs internet).
# Using the system default font (systemfonts, which has a bold face) makes the
# bold headings work and renders all glyphs correctly.
showtext::showtext_auto(FALSE)   # ensure showtext is not intercepting text
plot_family <- ""                # "" = graphics-device / systemfonts default

# To use a specific font instead, install it system-wide and set e.g.
# plot_family <- "Source Sans 3"  (it must be visible to systemfonts::system_fonts()).
```

# Load and prepare data
```{r}
# The results now come from two sources, kept separate upstream so the
# exploratory bias probes cannot leak into the pre-specified results:
#   - cox_ten_year_main_HR.csv         : PRE-SPECIFIED. Main models (M1–M5) and
#       effect modification (sex/age/education pre-specified; income/occupation
#       exploratory), written by 26_.
#   - ten_yr_biasprobe_interaction_HR.csv : EXPLORATORY bias probes (number of
#       assessments + lifestyle, HPA-only), written by 26b_.
df_main <-
  read.csv(here::here(
    "..", "results", "models", "cox_ten_year_main_HR.csv"
  ))

df_biasprobe <-
  read.csv(here::here(
    "..", "results", "models", "ten_yr_biasprobe_interaction_HR.csv"
  ))

df_forrest_plot <- dplyr::bind_rows(df_main, df_biasprobe)

# Data preparation and sorting
df_forrest_plot <- df_forrest_plot |>
  # Filter out the redundant 4-level occupation interaction (kept as detailed
  # SSYK, exposure m11, instead)
  filter(!exposure %in% c("m8")) |>

  # Create participant/event summary columns
  mutate(
    ctrl = glue("{comma(n_Control)} ({comma(events_Control)})"),
    hpa = glue("{comma(n_HPA)} ({comma(events_HPA)})")
  ) |>

  # Add header rows
  add_row(label = "Sex", block = "Sociodemographics", ctrl = " ", hpa = " ") |>
  add_row(
    label = "Age group",
    block = "Sociodemographics",
    ctrl = " ",
    hpa = " "
  ) |>
  add_row(
    label = "Education",
    block = "Sociodemographics",
    ctrl = " ",
    hpa = " "
  ) |>
  add_row(
    label = "Occupation",
    block = "Sociodemographics",
    ctrl = " ",
    hpa = " "
  ) |>
  add_row(
    label = "Income (% of median)",
    block = "Sociodemographics",
    ctrl = " ",
    hpa = " "
  ) |>
  add_row(label = "BMI", block = "Lifestyle related", ctrl = " ", hpa = " ") |>
  add_row(
    label = "VO₂max (mL·kg⁻¹·min⁻¹)",
    block = "Lifestyle related",
    ctrl = " ",
    hpa = " "
  ) |>
  add_row(
    label = "Exercise frequency",
    block = "Lifestyle related",
    ctrl = " ",
    hpa = " "
  ) |>
  add_row(
    label = "Health",
    block = "Lifestyle related",
    ctrl = " ",
    hpa = " "
  ) |>

  # Clean up labels
  mutate(
    label = case_when(
      label == "M1: birth cohort" ~ "M1: birth cohort",
      label == "M2: + age, sex, comorbidity & CVD" ~ "M2: + age, sex, comorbidity, prior CVD",
      label == "M3: + income" ~ "Age, sex, birth cohort,  \ncomorbidity, prior CVD & income",
      label == "M4: + occupation" ~ "M4: + income, occupation",
      label == "M5: + municipality & origin" ~ "M5: + municipality, origin",
      
      label == "HPA, 1 test" ~ "1 assessment",
      label == "HPA, 2 tests" ~ "2 assessments",
      label == "HPA, ≥3 tests" ~ "≥3 assessments",

      label == "BMI: Underweight" ~ "< 18.5",
      label == "BMI: Normal weight" ~ "18.5-25",
      label == "BMI: Overweight" ~ "25-30",
      label == "BMI: Obesity" ~ "≥30",

      label == "VO2max: Very high" ~ "≥50",
      label == "VO2max: High" ~ "40–49",
      label == "VO2max: Moderate" ~ "30–39",
      label == "VO2max: Low" ~ "20–29",
      label == "VO2max: Very low" ~ "<20",

      label == "Exercise: Good" ~ "3 ≥ times/week",
      label == "Exercise: Moderate" ~ "1-2 times/week",
      label == "Exercise: Low" ~ "Never/sometime",

      label == "Health: Good" ~ "Good or very good",
      label == "Health: Moderate" ~ "Neither good nor poor",
      label == "Health: Low" ~ "Poor or very poor",

      label == "Education" & exposure == "m11" ~ "Education professionals",

      TRUE ~ label
    ),

    # Set factor levels for proper ordering
    label = factor(
      label,
      levels = c(
        # Main models
        "M1: birth cohort",
        "M2: + age, sex, comorbidity, prior CVD",
        "Age, sex, birth cohort,  \ncomorbidity, prior CVD & income",
        "M4: + income, occupation",
        "M5: + municipality, origin",

        # Number of tests
        "≥3 assessments",
        "2 assessments",
        "1 assessment",

        # Sociodemographics
        "Sex",
        "Women",
        "Men",
        "Age group",
        "<35 yrs",
        "35–49 yrs",
        "≥50 yrs",
        "Education",
        "Tertiary",
        "Secondary",
        "Primary",
        "Occupation",
        "Managers",
        "Science and engineering",
        "Other professionals",
        "Education professionals",
        "Health care",
        "Associate professionals",
        "Administration and customer service",
        "Service and shop sales",
        "Personal care",
        "Manufacturing",
        "Mechanical manufacturing",
        "Building",
        "Transport",
        "Agriculture and forestry",
        "Cleaners",
        "Other elementary occupations",
        "Income (% of median)",
        "≥200%",
        "120–199%",
        "80–119%",
        "60–79%",
        "<60%",

        # Lifestyle related
        "BMI",
        "<18.5",
        "18.5-25",
        "25-30",
        "≥30",
        "VO₂max (mL·kg⁻¹·min⁻¹)",
        "≥50",
        "40–49",
        "30–39",
        "20–29",
        "<20",
        "Exercise frequency",
        "3 ≥ times/week",
        "1-2 times/week",
        "Never/sometime",
        "Health",
        "Good or very good",
        "Neither good nor poor",
        "Poor or very poor"
      )
    ),

    # Reassign blocks for the pre-specified vs exploratory framing.
    #  Pre-specified : the primary model (M3) + effect modification (sex/age/education)
    #  Exploratory   : the adjustment ladder (M1,M2,M4,M5), socioeconomic/occupation
    #                  effect modification, number of HPAs, and lifestyle bias probes
    block = case_when(
      exposure == "m3" ~ "Primary model (M3)",
      exposure %in% c("m1", "m2", "m4", "m5") ~ "Adjustment sets",
      exposure %in% c("m6", "m7", "m10") ~ "Effect modification",
      label %in% c("Sex", "Age group", "Education") ~ "Effect modification",
      exposure %in% c("m9", "m11") ~ "Socioeconomic",
      label %in% c("Occupation", "Income (% of median)") ~ "Socioeconomic",
      block == "Number of tests" ~ "HPAs within 5 yrs",
      block == "Lifestyle related" ~ "Lifestyle related*",
      TRUE ~ block
    ),

    # top-level tier used to split / label the figures
    tier = if_else(
      block %in% c("Primary model (M3)", "Effect modification"),
      "Pre-specified", "Exploratory"
    ),

    block = factor(
      block,
      levels = c(
        "Primary model (M3)",
        "Effect modification",
        "Adjustment sets",
        "Socioeconomic",
        "HPAs within 5 yrs",
        "Lifestyle related*"
      )
    ),
    tier = factor(tier, levels = c("Pre-specified", "Exploratory")),

    # colour key (all main/adjustment models share one colour)
    group_color = case_when(
      exposure %in% c("m1", "m2", "m3", "m4", "m5") ~ "Main models",
      block == "HPAs within 5 yrs" ~ as.character(block),
      exposure == "m6" ~ "Sex",
      exposure == "m7" ~ "Age",
      exposure == "m10" ~ "Education",
      exposure == "m11" ~ "Occupation",
      exposure == "m9" ~ "Income",
      block == "Lifestyle related*" ~ exposure,
    )
  ) |>

  # order by tier, then block, then label
  arrange(tier, block, label) |>
  mutate(
    row = row_number(),
    across(
      where(is.character),
      ~ .x |> str_replace_all("\\p{Z}", " ") |> str_squish()
    )
  )
```

# Forest-plot builder and outputs

We produce three figures from the same data:

- **combined** — one figure with two labelled tiers (Pre-specified / Exploratory).
- **main** — pre-specified only: the primary model **M3** + effect modification
  by sex, age and education.
- **supplementary** — everything secondary: the adjustment ladder (M1, M2, M4, M5),
  socioeconomic/occupation effect modification, number of HPAs within 5 years, and
  lifestyle bias probes.

```{r}
# colour palette (keyed on the group_color values)
my_cols <- c(
  "Main models"       = "#418286",
  "HPAs within 5 yrs" = "#e1800f",
  "Sex"               = "#418286",
  "Age"               = "#e1800f",
  "Education"         = "#418286",
  "Occupation"       = "#e1800f",
  "Income"            = "#418286",
  "BMI"               = "#e1800f",
  "Exercise"          = "#418286",
  "Health"            = "#e1800f",
  "VO2"               = "#418286"
)

# replace unicode glyphs that don't render reliably. NB: keep the real ≥ glyph —
# writing ">=" makes marquee treat a leading ">" as a blockquote and mangles the
# label. Only the VO2max sub/superscripts and the en dash need swapping.
fix_glyphs <- function(x) {
  x |>
    str_replace_all(fixed("VO₂max (mL·kg⁻¹·min⁻¹)"),
                    "VO2max (mL/kg/min)") |>
    str_replace_all(fixed("VO₂max"), "VO2max") |>
    str_replace_all("–", "-")       # en dash -> hyphen
}

# wrap heading text in marquee bold markdown (ckbplotr renders row labels with
# marquee::element_marquee, so **...** renders bold)
bold_md <- function(x) paste0("**", x, "**")

# build one forest plot from a (possibly filtered) slice of df_forrest_plot.
# row_levels: c("tier","group","label") for combined, c("group","label") for splits.
# (No in-plot caption — the HPA-only footnote goes in the manuscript figure legend.)
build_forest <- function(dat, row_levels) {
  results <- dat |>
    transmute(
      subgroup = as.character(label),
      ctrl, hpa,
      est = estimate, lci = conf.low, uci = conf.high,
      exposure, block, group_color
    ) |>
    drop_na(subgroup) |>
    # bold the sub-header rows (those with no estimate) and fix glyphs
    mutate(
      subgroup = if_else(is.na(est), bold_md(fix_glyphs(subgroup)), fix_glyphs(subgroup))
    )

  labels <- dat |>
    transmute(
      row,
      subgroup = as.character(label),
      group    = as.character(block),
      tier     = as.character(tier),
      est      = estimate
    ) |>
    distinct() |>
    mutate(
      # match the bolded/fixed subgroup in `results`, and bold the block/tier headings
      subgroup = if_else(is.na(est), bold_md(fix_glyphs(subgroup)), fix_glyphs(subgroup)),
      label = subgroup,
      group = bold_md(fix_glyphs(group)),
      tier  = bold_md(tier)
    ) |>
    select(-est) |>
    arrange(row) |>
    drop_na(label)

  # build the ggplot layers appended after the forest plot; caption is optional
  end_layers <- list(
    scale_x_log10(
      breaks = c(0.4, 0.6, 0.8, 1, 1.2, 1.5),
      labels = c("0.4", "0.6", "0.8", "1", "1.2", "1.5"),
      expand = expansion(mult = 0),
      sec.axis = dup_axis(
        name = NULL,
        breaks = c(.6, 0.8, 1, 1.2),
        labels = c("0.6", "0.8", "1", "1.2")
      )
    ),
    scale_colour_manual(values = my_cols, name = "Exposure\nblock"),
    scale_fill_manual(values = my_cols, guide = "none"),
    theme(
      axis.title.x.bottom = element_blank(),
      axis.title.x.top = element_text(face = "bold", vjust = 1),
      axis.text.x.top = element_text(),
      axis.line.x.top = element_blank(),
      axis.ticks.length.x.top = unit(0, "mm"),
      axis.ticks.x.top = element_line(colour = "black"),
      panel.background = element_rect(fill = "white", color = "white"),
      plot.background = element_rect(fill = "white", color = "white"),
      text = element_text(family = plot_family),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.caption = element_text(
        hjust = 0, size = 9, face = "italic", margin = margin(t = 5)
      )
    )
  )

  forest_plot(
    results,
    col.key = "subgroup",
    row.labels = labels,
    row.labels.levels = row_levels,
    col.lci = "lci",
    col.uci = "uci",
    exponentiate = FALSE,
    col.left.pos = unit(c(1.2, 31), "mm"),
    left.space = unit(60, "mm"),
    col.right.pos = unit(1.2, "mm"),
    right.space = unit(32.7, "mm"),
    xlab = "HPA vs controls\nHazard ratio (95% CI)",
    xlim = c(0.4, 1.5),
    xticks = c(0.6, 0.8, 1, 1.2),
    nullval = 1,
    col.left = c("ctrl", "hpa"),
    col.left.heading = c("Control\nN (events)", "HPA\nN (events)"),
    col.right.heading = "Hazard ratio\n(95% CI)",
    shape = 15,
    pointsize = 2.5,
    plotcolour = "grey20",
    colour = "group_color",
    cicolour = "group_color",
    fill = "#e1800f",
    plot.margin = margin(8, 8, 2, 8, "mm"),
    addarg = list(nullline = c("linetype = 'solid'", "colour = 'grey20'")),
    add = list(
      start = list(
        geom_vline(
          xintercept = c(0.6, 0.8, 1.2),
          linetype = "dashed", colour = "grey", size = 0.3
        )
      ),
      end = end_layers
    )
  )$plot  # forest_plot() returns list(plot, code); keep the ggplot
}

# Save each figure as PNG (raster) + SVG (vector). We use the ragg and svglite
# devices because ckbplotr renders row labels with marquee, which needs a device
# that supports the glyph API for the **bold** headings. The cairo_pdf device is
# broken on this machine ("failed to load cairo DLL") and the base pdf() device
# does not render marquee bold, so we do NOT write a PDF here. If a PDF is
# required, open the SVG in Illustrator/Affinity/Inkscape and export to PDF.
save_forest <- function(p, stub, height = 280) {
  png_dev <- if (requireNamespace("ragg", quietly = TRUE)) ragg::agg_png else "png"
  ggsave(
    here::here("..", "results", "figures", paste0(stub, ".png")),
    plot = p, width = 210, height = height, units = "mm", dpi = 300,
    device = png_dev
  )
  if (requireNamespace("svglite", quietly = TRUE)) {
    ggsave(
      here::here("..", "results", "figures", paste0(stub, ".svg")),
      plot = p, width = 210, height = height, units = "mm",
      device = svglite::svglite
    )
  } else {
    message("svglite not installed — SVG skipped for ", stub, " (PNG saved).")
  }
}
```

## Combined figure (two tiers)
```{r}
#| fig.height: 14
#| fig.width: 8
p_combined <- build_forest(
  df_forrest_plot,
  row_levels = c("tier", "group", "label")
)
p_combined
save_forest(p_combined, "forest_plot_combined", height = 320)
```

## Main figure (pre-specified only: M3 + effect modification)
```{r}
#| fig.height: 5
#| fig.width: 8
p_main <- build_forest(
  df_forrest_plot |> filter(tier == "Pre-specified") |> mutate(row = row_number()),
  row_levels = c("group", "label")
)
p_main
save_forest(p_main, "forest_plot_main", height = 130)
```

## Supplementary figure (adjustment ladder + exploratory)
```{r}
#| fig.height: 11
#| fig.width: 8
p_supp <- build_forest(
  df_forrest_plot |> filter(tier == "Exploratory") |> mutate(row = row_number()),
  row_levels = c("group", "label")
)
p_supp
save_forest(p_supp, "forest_plot_supplementary", height = 250)
```
