---
title: ""
format:
  html:
    embed-resources: true
    page-layout: full
    grid:
      margin-width: 100px
author: Daniel Väisänen
editor: source
toc: true
execute:
  warning: false
  echo: true
  eval: false
---

# Description

Immortal-time-bias check for the **number of HPAs** dose–response.

A participant cannot be in the "≥3 assessments" group without surviving long
enough to accrue three assessments — so classifying assessment count at baseline
(fixed) grants the higher-count groups "immortal" person-time and can spuriously
make more assessments look protective. Here we re-estimate the association with
**assessment count as a time-varying exposure** that increments at each
assessment date (a person only enters the "≥3" state *after* their third
assessment). We report the time-varying estimate next to the naïve
baseline-classified estimate; the gap between them is the immortal-time inflation.

**Estimand / control handling (Option A):** the 1:4 matched clusters are kept
intact. HPA participants carry the **time-varying** assessment count; each control
carries its matched participant's **baseline-assigned** category as a *constant*
(as in the reported dose–response). Set `control_mode <- "zero"` for Option B
(all controls fixed at the lowest category) if a different estimand is wanted.


# Libraries
```{r}
library(tidyverse)
library(survival)   # tmerge(), cumtdc(), coxph(), cox.zph()
library(broom)
```


# Data
```{r}
source(here::here("script", "01_data_source_file.R"))
source(here::here("script", "02_functions_source_file.R"))

control_mode <- "baseline"   # "baseline" = Option A, "zero" = Option B
```


# 1. Every assessment date (raw HPA file)
```{r}
# The saved df_n_tests only kept first/second/last dates, so we read all
# assessment dates from the raw cleaned HPA file (LopNr + Performed).
raw_hpa_path <- here::here(
  "..", "..", "..", "LE", "Data_och_analyser", "Data",
  "EEB_hpb_clean_2023-10-16.csv"
)
stopifnot("raw HPA file not found — check the path" = file.exists(raw_hpa_path))

all_assessments <- readr::read_csv(
  raw_hpa_path,
  col_select = c(LopNr, Performed),
  show_col_types = FALSE
) |>
  mutate(Performed = as.Date(Performed)) |>
  filter(!is.na(Performed)) |>
  distinct(LopNr, Performed)
```


# 2. Time zero, follow-up, baseline category, assessment times
```{r}
# One row per person in the analytic cohort. Time zero is HPA_date (the survival
# baseline that risk_time_art_10yrs is measured from). risk_time_art_10yrs is the
# follow-up length in days; risk_art_flag_10yrs the event indicator.
base <- data_cvd |>
  transmute(
    LopNr,
    id_cluster,
    treated,
    time_zero = as.Date(HPA_date),
    tstop     = as.numeric(risk_time_art_10yrs),
    event     = as.integer(risk_art_flag_10yrs),
    n_tests_within_5,
    baseline_cat = n_tests_within_5_fac_vs_ctrl,   # controls carry the matched HPA's category (imputed in 01_)
    Age, Sex, birth_cohort, comorbidity, cvd_art_before_HPA_flag, IncomeLevel_CSFVI
  ) |>
  # recode the baseline category to clean, shared levels
  mutate(
    baseline_cat = fct_recode(
      baseline_cat,
      "1 assessment"  = "HPA, 1 test",
      "2 assessments" = "HPA, 2 tests",
      "≥3 assessments" = "HPA, ≥3 tests"
    ),
    baseline_cat = factor(
      baseline_cat,
      levels = c("1 assessment", "2 assessments", "≥3 assessments")
    )
  ) |>
  # tmerge requires strictly positive follow-up
  filter(!is.na(tstop), tstop > 0, !is.na(event))

# assessment times (days since time zero) for HPA participants, within the
# 5-year window and AT OR AFTER time zero.
assess_long <- base |>
  filter(treated == "HPA") |>
  select(LopNr, time_zero, tstop) |>
  inner_join(all_assessments, by = "LopNr") |>
  mutate(atime = as.numeric(Performed - time_zero)) |>          # days since time zero
  filter(atime >= 0, atime <= 1826) |>                          # 5-year window (1826 days)
  arrange(LopNr, atime)

# sanity: assessment count from raw dates should track n_tests_within_5
assess_counts <- assess_long |> count(LopNr, name = "n_assess_raw")
base |>
  filter(treated == "HPA") |>
  left_join(assess_counts, by = "LopNr") |>
  mutate(n_assess_raw = replace_na(n_assess_raw, 0L)) |>
  summarise(
    mean_saved = mean(n_tests_within_5, na.rm = TRUE),
    mean_raw   = mean(n_assess_raw),
    agree_pct  = mean(n_tests_within_5 == n_assess_raw, na.rm = TRUE) * 100
  )
```


# 3. Data check 1 — no assessment before time zero
```{r}
# Any negative atime would silently break the tmerge split. We already filtered
# atime >= 0 above; here we confirm none were dropped for being pre-baseline.
pre_baseline <- base |>
  filter(treated == "HPA") |>
  select(LopNr, time_zero) |>
  inner_join(all_assessments, by = "LopNr") |>
  filter(as.numeric(Performed - time_zero) < 0,
         as.numeric(Performed - time_zero) >= -1826)   # within window but before t0

cat("Assessments dated before time zero (dropped):", nrow(pre_baseline), "\n")
```


# 3b. Event timing relative to later assessments (immortal time made concrete)

How often did a participant's CVD event occur **before** the assessment that
defines their higher baseline count? Those are exactly the events the
baseline-classified ("naïve") model credits to the higher-count group but the
time-varying model reattributes to the lower count — the immortal time, counted
directly. (Incident CVD here is mostly non-fatal, so a participant can have an
event and then return for a later assessment.)

```{r}
# rank each HPA participant's assessments: 1 = baseline HPA (atime ~0), 2 = second, 3 = third
assess_ranked <- assess_long |>
  arrange(LopNr, atime) |>
  mutate(assess_no = row_number(), .by = LopNr)

# second / third assessment times (days since time zero), one row per person
assess_wide <- assess_ranked |>
  filter(assess_no %in% c(2, 3)) |>
  select(LopNr, assess_no, atime) |>
  tidyr::pivot_wider(
    names_from = assess_no, values_from = atime, names_prefix = "t_assess_"
  )

# join each HPA participant's event time to their 2nd/3rd assessment times
hpa_events <- base |>
  filter(treated == "HPA") |>
  select(LopNr, tstop, event) |>
  left_join(assess_wide, by = "LopNr")

# denominators are people who actually reached 2 (3) assessments within 5 years
among_2 <- hpa_events |> filter(!is.na(t_assess_2))
among_3 <- hpa_events |> filter(!is.na(t_assess_3))

event_timing <- tibble(
  group = c("Reached 2 assessments", "Reached 3 assessments"),
  n = c(nrow(among_2), nrow(among_3)),
  cvd_events = c(sum(among_2$event), sum(among_3$event)),
  # events occurring before the 2nd / 3rd assessment date
  cvd_before_that_assessment = c(
    sum(among_2$event == 1 & among_2$tstop < among_2$t_assess_2),
    sum(among_3$event == 1 & among_3$tstop < among_3$t_assess_3)
  )
) |>
  mutate(
    pct_of_events = round(100 * cvd_before_that_assessment / cvd_events, 1),
    pct_of_people = round(100 * cvd_before_that_assessment / n, 2)
  )

print(event_timing)

write_csv(
  event_timing,
  here::here("..", "results", "tables", "immortal_time_event_timing.csv")
)

# ready-to-paste Table S1 footnote, numbers filled straight from the counts above
r3 <- event_timing |> filter(group == "Reached 3 assessments")
footnote_s1 <- sprintf(
  paste0(
    "Of the %s CVD events among participants who completed three or more ",
    "assessments within five years, %s (%.1f%%) occurred before the third ",
    "assessment; the baseline-classified analysis credits these events to the ",
    "≥3 group, whereas the time-varying analysis attributes them, and the ",
    "person-time preceding each assessment, to the lower count."
  ),
  format(r3$cvd_events, big.mark = ","),
  format(r3$cvd_before_that_assessment, big.mark = ","),
  r3$pct_of_events
)
cat("\nTable S1 footnote:\n", footnote_s1, "\n")
```


# 4. Counting-process dataset via tmerge (time-varying n_hpa)
```{r}
# Post-baseline assessments only (the first assessment is time zero → baseline
# count of 1). Keep those strictly inside follow-up so no split lands on/after tstop.
assess_increments <- assess_long |>
  filter(atime > 0, atime < tstop) |>
  select(LopNr, atime)

# cvd = event() sets the terminal interval; cumtdc() counts post-baseline
# assessments accrued so far.
cp <- tmerge(base, base, id = LopNr, cvd = event(tstop, event))
cp <- tmerge(cp, assess_increments, id = LopNr, n_post = cumtdc(atime))

cp <- cp |>
  mutate(
    n_post = replace_na(as.integer(n_post), 0L),
    # baseline count is 1; each post-baseline assessment adds 1; cap at 3
    n_hpa_count = pmin(1L + n_post, 3L),
    n_hpa_tv = if (control_mode == "zero") {
      # Option B: controls fixed at the lowest category
      factor(
        case_when(
          treated == "Control" ~ "1 assessment",
          n_hpa_count == 1 ~ "1 assessment",
          n_hpa_count == 2 ~ "2 assessments",
          n_hpa_count >= 3 ~ "≥3 assessments"
        ),
        levels = c("1 assessment", "2 assessments", "≥3 assessments")
      )
    } else {
      # Option A: controls carry their matched baseline category (constant)
      factor(
        case_when(
          treated == "Control" ~ as.character(baseline_cat),
          n_hpa_count == 1 ~ "1 assessment",
          n_hpa_count == 2 ~ "2 assessments",
          n_hpa_count >= 3 ~ "≥3 assessments"
        ),
        levels = c("1 assessment", "2 assessments", "≥3 assessments")
      )
    }
  )
```


# 5. Data check 2 — events and person-time preserved after splitting
```{r}
checks <- tibble(
  events_before      = sum(base$event),
  events_after       = sum(cp$cvd),
  events_ok          = sum(base$event) == sum(cp$cvd),
  persontime_before  = sum(base$tstop),
  persontime_after   = sum(cp$tstop - cp$tstart),
  # allow a tiny numeric tolerance
  persontime_ok      = abs(sum(base$tstop) - sum(cp$tstop - cp$tstart)) < 1,
  bad_intervals      = sum(cp$tstop <= cp$tstart),   # must be 0
  any_na_interval    = any(is.na(cp$tstart) | is.na(cp$tstop))
)
checks

stopifnot(
  "events lost/duplicated at split boundaries" = checks$events_ok,
  "person-time not preserved after splitting"  = checks$persontime_ok,
  "invalid (tstop <= tstart) intervals present" = checks$bad_intervals == 0
)
```


# 6. Estimate: naïve vs time-varying assessment count

The **HPA-only** contrast is the primary analysis: the immortal time lives entirely
within the HPA arm (you cannot be "≥3 assessments" without surviving to accrue
three), so restricting to HPA participants with "1 assessment" as the reference
isolates the bias most cleanly. The **matched 1:4** version is reported as a
secondary estimand (Option A); because each control inherits its matched
participant's category, its coefficients are near-null and less informative for the
dose–response question.

```{r}
cov_rhs <- "Age + Sex + birth_cohort + comorbidity + cvd_art_before_HPA_flag + IncomeLevel_CSFVI"

tidy_hr <- function(fit, which, panel) {
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) |>
    filter(str_detect(term, "assessment")) |>
    transmute(
      panel, model = which,
      term = str_remove(term, "^(baseline_cat|n_hpa_tv)"),
      estimate, conf.low, conf.high, p.value
    )
}

## PRIMARY — HPA participants only -----------------------------------------
base_hpa <- base |> filter(treated == "HPA")
cp_hpa   <- cp   |> filter(treated == "HPA")

# NAÏVE: assessment count fixed at baseline (the reported dose–response; this is
# the estimate that immortal time inflates).
fit_naive_hpa <- coxph(
  as.formula(paste("Surv(tstop, event) ~ baseline_cat +", cov_rhs)),
  data = base_hpa
)
# TIME-VARYING: count increments over follow-up (immortal time removed).
fit_tv_hpa <- coxph(
  as.formula(paste("Surv(tstart, tstop, cvd) ~ n_hpa_tv +", cov_rhs)),
  data = cp_hpa
)

## SECONDARY — matched 1:4 (Option A), clusters intact ---------------------
fit_naive_matched <- coxph(
  as.formula(paste("Surv(tstop, event) ~ baseline_cat +", cov_rhs)),
  data = base, cluster = id_cluster
)
fit_tv_matched <- coxph(
  as.formula(paste("Surv(tstart, tstop, cvd) ~ n_hpa_tv +", cov_rhs)),
  data = cp, cluster = id_cluster
)

comparison <- bind_rows(
  tidy_hr(fit_naive_hpa,     "Naïve (baseline count)",   "HPA-only (primary)"),
  tidy_hr(fit_tv_hpa,        "Time-varying (corrected)", "HPA-only (primary)"),
  tidy_hr(fit_naive_matched, "Naïve (baseline count)",   "Matched 1:4 (secondary)"),
  tidy_hr(fit_tv_matched,    "Time-varying (corrected)", "Matched 1:4 (secondary)")
) |>
  mutate(`HR (95% CI)` = sprintf("%.2f (%.2f–%.2f)", estimate, conf.low, conf.high))

comparison |> arrange(panel, term, model) |> print(n = 50)

write_csv(
  comparison,
  here::here("..", "results", "tables", "immortal_time_ntests_comparison.csv")
)
```


# 7. Proportional-hazards test (HPA-only time-varying term)
```{r}
# Table only. The Schoenfeld residual plot is one point per event (tens of
# thousands): hard to read and memory-heavy in a batch run, so it is omitted.
# A non-significant global p means PH holds and the constant-HR summary is
# adequate; a significant p for the ≥3 term would itself indicate WHEN the
# count/CVD divergence emerges.
zph_hpa <- cox.zph(fit_tv_hpa)
zph_hpa
```


# 8. Printable report + manuscript summary
```{r}
# headline contrast: ≥3 vs 1 assessment, HPA participants
row_naive <- comparison |>
  filter(
    panel == "HPA-only (primary)",
    term == "≥3 assessments",
    str_detect(model, "Naïve")
  )
row_tv <- comparison |>
  filter(
    panel == "HPA-only (primary)",
    term == "≥3 assessments",
    str_detect(model, "Time-varying")
  )

pct_removed <- (log(row_naive$estimate) - log(row_tv$estimate)) /
  log(row_naive$estimate) *
  100
tv_sig <- row_tv$conf.low > 1 | row_tv$conf.high < 1
global_p <- tryCatch(zph_hpa$table["GLOBAL", "p"], error = function(e) NA_real_)

# side-by-side markdown table: naïve vs time-varying, HPA participants (ref = 1)
hr_str <- function(term_val, model_pat) {
  r <- comparison |>
    filter(
      panel == "HPA-only (primary)",
      term == term_val,
      str_detect(model, model_pat)
    )
  if (nrow(r) == 0) "—" else r$`HR (95% CI)`[1]
}
result_tbl <- c(
  "| Number of HPAs (vs 1 assessment) | Naïve HR (95% CI) | Time-varying HR (95% CI) |",
  "|---|---|---|",
  sprintf(
    "| ≥3 assessments | %s | %s |",
    hr_str("≥3 assessments", "Naïve"),
    hr_str("≥3 assessments", "Time-varying")
  ),
  sprintf(
    "| 2 assessments | %s | %s |",
    hr_str("2 assessments", "Naïve"),
    hr_str("2 assessments", "Time-varying")
  )
)

md <- c(
  "# Immortal-time-bias check — number of HPAs",
  "",
  "## What was done",
  "The number of HPAs within 5 years (1, 2, ≥3) was re-estimated as a **time-varying**",
  "exposure with a counting-process Cox model (`survival::tmerge` + `cumtdc`): a",
  "participant enters the '≥3' state only *after* their third assessment date, so the",
  "person-time accrued before that is no longer credited to the higher-count group.",
  "This removes immortal-time bias. Primary analysis: HPA participants only,",
  "'1 assessment' as reference, adjusted for the primary-model (M3) covariates.",
  "",
  "## Results (HPA participants, reference = 1 assessment)",
  "",
  result_tbl,
  "",
  "## Headline result — ≥3 vs 1 assessment (HPA participants)",
  sprintf(
    "- Naïve (count fixed at baseline): **%s**, p = %.3g",
    row_naive$`HR (95% CI)`,
    row_naive$p.value
  ),
  sprintf(
    "- Time-varying (immortal-time corrected): **%s**, p = %.3g",
    row_tv$`HR (95% CI)`,
    row_tv$p.value
  ),
  sprintf(
    "- Immortal time accounted for ~%.0f%% of the apparent (log-scale) association; the corrected estimate is %s significant at p < 0.05.",
    pct_removed,
    if (tv_sig) "still" else "no longer"
  ),
  sprintf(
    "- Proportional-hazards test (time-varying term): global p = %.3g.",
    global_p
  ),
  "",
  "## Interpretation",
  "Correcting for immortal time attenuates the apparent protective dose–response of",
  "more assessments. Removing immortal time does **not** remove confounding —",
  "participants who return for more assessments may be healthier or more",
  "health-conscious — so the residual association cannot be interpreted causally.",
  "This supports treating the number-of-HPAs analysis as exploratory rather than as",
  "evidence of a causal dose–response."
)

writeLines(md)
writeLines(
  md,
  here::here("..", "results", "tables", "immortal_time_report.md.md")
)
```
